\newpage

%\section*{Publication list}
%\vspace*{-1cm}

%\subsection*{Book chapters}
%\subsection*{Journal papers}
%\subsection*{Peer-reviewed conference papers}
%\subsection*{Abstracts}
%\subsection*{Preprints}
%\subsection*{Theses}

\begin{thebibliography}{100}

\subsection*{Book chapters}

\bibitem{inbook:15}
\underline{A.~Roche}.
\newblock {\em {Image, Video \& 3D Data Registration: Medical, Satellite \&
  Video Processing Applications with Quality Metrics}}, chapter {Medical Image
  Registration Measures}.
\newblock Wiley, August 2015.

\bibitem{book:10}
J.-B. Poline, B.~Thirion, \underline{A.~Roche}, and S.~M\'eriaux.
\newblock {\em Foundational Issues in Human Brain Mapping}, chapter
  {Intersubject variability in fMRI data: causes, consequences, and related
  analysis strategies}, pages 173--192.
\newblock MIT press, 2010.

\bibitem{book:05}
X.~Pennec, N.~Ayache, \underline{A.~Roche}, and P.~Cachier.
\newblock {\em Multi-Sensor Image Fusion and Its Applications}, chapter
  Non-rigid {MR/US} registration for tracking brain deformations, pages
  107--143.
\newblock CRC Press, Boca Raton, USA, first edition, 2006.

\bibitem{book:02}
S.~Warfield, A.~Guimond, \underline{A.~Roche}, A.~Bharatha, A.~Tei, F.~Talos, J.~Rexilius,
  J.~Ruiz-Alzola, C.-F. Westin, S.~Haker, S.~Angenent, A.~Tannenbaum,
  F.~Jolesz, and R.~Kikinis.
\newblock {\em Brain Mapping: The Methods}, chapter Advanced Nonrigid
  Registration Algorithms for Image Fusion, pages 661--690.
\newblock Academic Press, San Diego, USA, second edition, 2002.


\subsection*{Journal papers}

\bibitem{flash:16}
\underline{A.~Roche}, B.~Mar\'echal, T.~Kober, G.~Krueger, P.~Hagmann, P.~Maeder, and
  R.~Meuli.
\newblock {Assessing brain volumes using MorphoBox prototype}.
\newblock {\em MAGNETOM Flash, the Siemens magazine of MRI: RSNA Edition},
  2017.

\bibitem{ijasp:13}
\underline{A.~Roche}.
\newblock {Approximate inference via variational sampling}.
\newblock {\em International Journal of Applied Statistics and Probability},
  1(3):110--120, 2013.

\bibitem{tmi:11}
\underline{A.~Roche}.
\newblock {A four-dimensional registration algorithm with application to joint
  correction of motion and slice timing in fMRI}.
\newblock {\em IEEE Transactions on Medical Imaging}, 30(8):1546--1554, 2011.

\bibitem{media:11}
\underline{A.~Roche}, D.~Ribes, M.~Bach~Cuadra, and G.~Krueger.
\newblock {On the Convergence of EM-Like Algorithms for Image Segmentation
  using Markov Random Fields}.
\newblock {\em Medical Image Analysis}, 15(6):830--839, 2011.

\bibitem{ni:07}
\underline{A.~Roche}, S.~M\'eriaux, M.~Keller, and B.~Thirion.
\newblock {Mixed-effects statistics for group analysis in fMRI: A nonparametric
  maximum likelihood approach}.
\newblock {\em Neuroimage}, 38:501--510, 2007.

\bibitem{tmi:01b}
\underline{A.~Roche}, X.~Pennec, G.~Malandain, and N.~Ayache.
\newblock {Rigid Registration of 3D Ultrasound with MR Images: a New Approach
  Combining Intensity and Gradient Information}.
\newblock {\em IEEE Transactions on Medical Imaging}, 20(10):1038--1049, 2001.

\bibitem{ijist:00}
\underline{A.~Roche}, G.~Malandain, and N.~Ayache.
\newblock {Unifying Maximum Likelihood Approaches in Medical Image
  Registration}.
\newblock {\em International Journal of Imaging Systems and Technology: Special
  issue on 3D imaging}, 11:71--80, 2000.

\bibitem{pv:16}
M.J. Fartaria, A.~Sorega, G.~Krueger, T.~Kober, C.~Granziera, M.~Bach~Cuadra,
  and \underline{A.~Roche}.
\newblock {Partial Volume-Aware Assessment of Multiple Sclerosis Lesions}.
\newblock 2018.
\newblock {\em NeuroImage:Clinical}.
\newblock Accepted.

\bibitem{plos:16}
G.~Bonnier, T.~Kober, M.~Schluep, R.~Du~Pasquier, G.~Krueger, R.~Meuli,
  C.~Granziera, and \underline{A.~Roche}.
\newblock {A New Approach for Deep Gray Matter Analysis Using Partial-Volume
  Estimation}.
\newblock {\em PlosOne}, 11(2):e0148631, February 2016.

\bibitem{plos:14}
K.~O'Brien, G.~Krueger, P.~Hagmann, P.~Maeder, T.~Kober, J.~Marques,
  F.~Lazeyras, and \underline{A.~Roche}.
\newblock {Robust T1-weighted structural brain imaging and morphometry at 7T
  using MP2RAGE}.
\newblock {\em PlosOne}, 9(6):e99676, June 2014.

\bibitem{neuroradiology:17}
J.~Boto, G.~Gkinis, \underline{A.~Roche}, T.~Kober, B.~Mar{\'e}chal, N.~Ortiz, K.-O.~L{\"o}vblad, F.~Lazeyras, and M.I.~Vargas.
\newblock {Evaluating anorexia-related brain atrophy using MP2RAGE-based morphometry}.
\newblock {\em European Radiology}, 27(12):5064--5072, 2017.

\bibitem{investigative_radiology:17}
M.J.~Fartaria, K.~O'Brien, A.~Sorega, G.~Bonnier, \underline{A.~Roche}, P.~Falkovskiy, G.~Krueger, T.~Kober, M.~Bach Cuadra, and C.~Granziera.
\newblock {An ultra-high field study of cerebellar pathology in early relapsing-remitting multiple sclerosis using MP2RAGE}.
\newblock {\em Investigative radiology}, 52(5):265--273, 2017.

\bibitem{ni:16}
P.~Falkovskiy, D.~Brenner, T.~Feiweier, S.~Kannengiesser, B.~Mar\'echal,
  T.~Kober, \underline{A.~Roche}, K.~Thostenson, R.~Meuli, D.~Reyes, T.~Stoecker, M.A.
  Bernstein, J.-P. Thiran, and G.~Krueger.
\newblock {Comparison of accelerated T1-weighted whole-brain structural-imaging
  protocols}.
\newblock {\em Neuroimage}, 124:157--167, 2016.

\bibitem{neurorad:16}
S.~Haller, P.~Falkovskiy, R.~Meuli, J.-P. Thiran, G.~Krueger, K.-O. Lovblad,
  \underline{A.~Roche}, T.~Kober, and B.~Mar\'echal.
\newblock {Basic MR sequence parameters systematically bias automated brain
  volume estimation}.
\newblock {\em Neuroradiology}, 2016.

\bibitem{jmri:15}
M.J. Fartaria, G.~Bonnier, \underline{A.~Roche}, T.~Kober, R.~Meuli, D.~Rotzinger,
  R.~Frackowiak, M.~Schluep, R.~Du~Pasquier, J.-P. Thiran, G.~Krueger,
  M.~Bach~Cuadra, and C.~Granziera.
\newblock Automated detection of white matter and cortical lesions in early
  stages of multiple sclerosis.
\newblock {\em Journal of Magnetic Resonance Imaging}, 43(6):1445--1454, 2015.

\bibitem{biomed:15}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.E. Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, T.J. Du~Pasquier, R.~Sumpf, J.~Frahm,
  J.-P. Thiran, G.~Krueger, and C.~Granziera.
\newblock {Multicontrast MRI quantification of focal inflammation and
  degeneration in multiple sclerosis}.
\newblock {\em BioMed research international}, 2015(Article ID 569123), 2015.

\bibitem{nic:15}
C.~Granziera, A.~Daducci, A.~Donati, G.~Bonnier, D.~Romascano, \underline{A.~Roche},
  M.~Bach~Cuadra, D.~Schmitter, S.~Kl{\"o}ppel, R.~Meuli, A.~von Gunten, and
  G.~Krueger.
\newblock {A multi-contrast MRI study of microstructural brain damage in
  patients with mild cognitive impairment}.
\newblock {\em NeuroImage: Clinical}, 8:631--639, 2015.

\bibitem{nic:14}
D.~Schmitter, \underline{A.~Roche}, B.~Mar\'echal, D.~Ribes, A.~Abdulkadir, M.~Bach~Cuadra,
  A.~Daducci, C.~Granziera, S.~Kl\"oppel, P.~Maeder, R.~Meuli, and G.~Krueger.
\newblock {An evaluation of volume-based morphometry for prediction of mild
  cognitive impairment and Alzheimer's disease}.
\newblock {\em NeuroImage: Clinical}, 7:7--17, 2015.

\bibitem{hbm:15}
D.~Romascano, D.E. Meskaldji, G.~Bonnier, S.~Simioni, D.~Rotzinger, Y.-C. Lin,
  G.~Menegaz, \underline{A.~Roche}, M.~Schluep, R.~Du~Pasquier, J.~Richiardi, D.~Van
  De~Ville, A.~Daducci, T.~Sumpf, J.~Fraham, J.-P. Thiran, G.~Krueger, and
  C.~Granziera.
\newblock {Multicontrast connectometry: A new tool to assess cerebellum
  alterations in early relapsing-remitting multiple sclerosis}.
\newblock {\em Human Brain Mapping}, 36(4):1609--1619, 2015.

\bibitem{actn:14}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.~Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, R.~Du~Pasquier, T.~Sumpf, J.~Frahm, J.-P.
  Thiran, G.~Krueger, and C.~Granziera.
\newblock {Advanced MRI unravels the nature of tissue alterations in early
  multiple sclerosis}.
\newblock {\em Annals of clinical and translational neurology}, 1(6):423--432,
  2014.

\bibitem{hbm:14}
C.~Granziera, A.~Daducci, D.~Romascano, \underline{A.~Roche}, G.~Helms, G.~Krueger, and
  N.~Hadjikhani.
\newblock {Structural abnormalities in the thalamus of migraineurs with aura: A
  multiparametric study at 3 T}.
\newblock {\em Human Brain Mapping}, 35(4):1461--1468, 2013.

\bibitem{cerebellum:13}
C.~Granziera, D.~Romascano, A.~Daducci, \underline{A.~Roche}, M.~Vincent, G.~Krueger, and
  N.~Hadjikhani.
\newblock {Migraineurs without aura show microstructural abnormalities in the
  cerebellum and frontal lobe}.
\newblock {\em The Cerebellum}, 12(6):812--818, 2013.

\bibitem{neurology:12}
C.~Granziera, A.~Daducci, D.~Meskaldji, \underline{A.~Roche}, P.~Maeder, P.~Michel,
  N.~Hadjikhani, G.~Sorensen, R.~Frackowiak, J.-P. Thiran, R.~Meuli, and
  G.~Krueger.
\newblock {A new early and automated MRI-based predictor of motor improvement}.
\newblock {\em Neurology}, 79(1):39--46, 2012.

\bibitem{sinica:08}
M.~Keller, \underline{A.~Roche}, A.~Tucholka, and B.~Thirion.
\newblock {Dealing with Spatial Normalization Errors in fMRI Group Inference
  using Hierarchical Modeling}.
\newblock {\em Statistica Sinica}, 18(4):1357--1374, 2008.

\bibitem{media:08}
C.~Poupon, \underline{A.~Roche}, J.~Dubois, J.-F. Mangin, and F.~Poupon.
\newblock {Real-Time MR Diffusion Tensor and Q-ball Imaging using Kalman
  Filtering}.
\newblock {\em Medical Image Analysis}, 12(5):527--534, 2008.
\newblock Elsevier Medical Image Analysis -- MICCAI'07 best paper.

\bibitem{tmi:07}
E.~Duchesnay, A.~Cachia, \underline{A.~Roche}, D.~Rivi\`ere, Y.~Cointepas,
  D.~Papadopoulos-Orfanos, M.~Zilbovicius, J.-L. Martinot, and J.-F. Mangin.
\newblock Classification from cortical folding patterns.
\newblock {\em IEEE Transactions on Medical Imaging}, 26(4):553--565, 2007.

\bibitem{ni:07b}
B.~Thirion, P.~Pinel, S.~M\'eriaux, \underline{A.~Roche}, S.~Dehaene, and J.-B. Poline.
\newblock {Analysis of a Large fMRI cohort: Statistical and Methodological
  Issues for Group Analyses}.
\newblock {\em Neuroimage}, 35(1):105--120, 2007.

\bibitem{tmi:07b}
B.~Thirion, P.~Pinel, A.~Tucholka, \underline{A.~Roche}, P.~Ciuciu, J.-F. Mangin, and J.-B.
  Poline.
\newblock {Structural Analysis of fMRI Data Revisited: Improving the
  Sensitivity and Reliability of fMRI Group Studies}.
\newblock {\em IEEE Transactions on Medical Imaging}, 26(9):1256--1269, 2007.

\bibitem{pnas:06}
G.~Dehaene-Lambertz, L.~Hertz-Pannier, J.~Dubois, S.~M\'eriaux, \underline{A.~Roche}, and
  M.~Sigman.
\newblock Functional organization of perisylvian activation during presentation
  of sentences in preverbal infants.
\newblock {\em Proceedings of the National Academy of Sciences},
  103:14240--14245, 2006.

\bibitem{hbm:06}
S.~M\'eriaux, \underline{A.~Roche}, G.~Dehaene-Lambertz, B.~Thirion, and J.-B. Poline.
\newblock {Combined permutation test and mixed-effect model for group average
  analysis in fMRI}.
\newblock {\em Human Brain Mapping}, 27(5):402--410, 2006.

\bibitem{hbm:06b}
B.~Thirion, G.~Flandin, P.~Pinel, \underline{A.~Roche}, and J.-B. Ciuciu, P.~Poline.
\newblock {Dealing with the shortcomings of spatial normalization:
  Multi-subject parcellation of fMRI datasets}.
\newblock {\em Human Brain Mapping}, 27(8):678--693, August 2006.

\bibitem{pls:05}
J.-B. Poline, A.~Ciuciu, \underline{A.~Roche}, and B.~Thirion.
\newblock Quelle confiance accorder aux images du cerveau en action?
\newblock {\em Pour la Science}, 338:138--142, 2005.

\bibitem{tmi:02}
L.~Freire, \underline{A.~Roche}, and J.-F. Mangin.
\newblock {What is the best similarity measure for motion correction in fMRI
  time series?}
\newblock {\em IEEE Transactions on Medical Imaging}, 21(5):470--484, 2002.

\bibitem{tmi:01}
A.~Guimond, \underline{A.~Roche}, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Brain Warping Using the Demons Algorithm and Adaptative
  Intensity Corrections}.
\newblock {\em IEEE Transactions on Medical Imaging}, 20(1):58--69, 2001.

\bibitem{ivc:00}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, X.~Pennec, and N.~Ayache.
\newblock {Reconstructing a 3D Structure from Serial Histological Sections}.
\newblock {\em Image and Vision Computing}, 19(1--2):25--31, 2000.


\subsection*{Peer-reviewed conference papers}

\bibitem{miccai:14}
\underline{A.~Roche} and F.~Forbes.
\newblock {Partial volume estimation in brain MRI revisited}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'14)}, volume 8673 of {\em Lecture Notes in Computer Science}, pages
  771--778, Boston, USA, September 2014. Springer Verlag.

\bibitem{miccai:12}
\underline{A.~Roche}.
\newblock {Closed-Form Relaxation for MRF-MAP Tissue Classification Using
  Discrete Laplace Equations}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'12)}, volume 7511 of {\em Lecture Notes in Computer Science}, pages
  355--362, Nice, France, October 2012. Springer Verlag.

\bibitem{isbi:04}
\underline{A.~Roche}, P.-J. Layahe, and J.-B. Poline.
\newblock {Incremental Activation Detection in fMRI Time Series using Kalman
  Filtering}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 376--379, Washington, DC, USA, April 2004.

\bibitem{miccai:04}
\underline{A.~Roche}, P.~Pinel, S.~Dehaene, and J.-B. Poline.
\newblock {Solving Incrementally the Fitting and Detection Problems in fMRI
  Time Series}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'04)}, volume 3217 of {\em Lecture Notes in Computer Science}, pages
  719--726, Saint-Malo, France, September 2004. Springer Verlag.

\bibitem{eccv:00}
\underline{A.~Roche}, A.~Guimond, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Elastic Matching of Brain Images}.
\newblock In {\em European Conference on Computer Vision (ECCV'00)}, volume
  1843 of {\em Lecture Notes in Computer Science}, pages 511--527, Dublin,
  Ireland, June 2000. Springer Verlag.

\bibitem{miccai:00}
\underline{A.~Roche}, X.~Pennec, M.~Rudolph, D.~P. Auer, G.~Malandain, S.~Ourselin, L.~M.
  Auer, and N.~Ayache.
\newblock {Generalized Correlation Ratio for Rigid Registration of 3D
  Ultrasound with MR Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 567--577, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{miccai:99}
\underline{A.~Roche}, G.~Malandain, N.~Ayache, and S.~Prima.
\newblock {Towards a Better Comprehension of Similarity Measures used in
  Medical Image Registration}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'99)}, volume 1679 of {\em Lecture Notes in Computer Science}, pages
  555--566, Cambridge, UK, September 1999. Springer Verlag.

\bibitem{miccai:98}
\underline{A.~Roche}, G.~Malandain, X.~Pennec, and N.~Ayache.
\newblock {The Correlation Ratio as a New Similarity Measure for Multimodal
  Image Registration}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'98)}, volume 1496 of {\em Lecture Notes in Computer Science}, pages
  1115--1124, Cambridge, USA, October 1998. Springer Verlag.

\bibitem{miccai:09}
M.~Keller, M.~Lavielle, M.~Perrot, and \underline{A.~Roche}.
\newblock {Anatomically Informed Bayesian Model Selection for fMRI Group Data
  Analysis}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'09)}, volume 5762 of {\em Lecture Notes in Computer Science}, pages
  450--457, London, UK, September 2009. Springer Verlag.

\bibitem{isbi:08}
M.~Keller and \underline{A.~Roche}.
\newblock {Increased Sensitivity in fMRI Group Analyses using Mixed-Effect
  Modeling}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'08)},
  pages 548--551, Paris, France, May 2008.

\bibitem{isbi:08b}
P.~Ciuciu, T.~Vincent, A.-L. Foulque, and \underline{A.~Roche}.
\newblock {Improved fMRI group studies based on spatially varying
  non-parametric BOLD signal modeling}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'08)},
  pages 1263--1266, Paris, France, May 2008.

\bibitem{isbi:07}
P.~Ciuciu, P.~Abry, C.~Rabrait, H.~Wendt, and \underline{A.~Roche}.
\newblock {Leader-Based Multifractal Analysis for Evi fmri Time Series: Ongoing
  vs Task-Related Brain Activity}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'07)},
  pages 404--407, Washington, DC, USA, May 2007.

\bibitem{miccai:01b}
S.~Granger, X.~Pennec, and \underline{A.~Roche}.
\newblock {Rigid Point-Surface Registration using an EM Variant of ICP for
  Computed Guided Oral Implantology}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 752--761, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miccai:08}
M.~Bhattacharjee, A.~Pitiot, \underline{A.~Roche}, D.~Dormont, and E.~Bardinet.
\newblock {Anatomy-Preserving Nonlinear Registration of Deep Brain ROIs using
  Confidence-Based Block-Matching}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'08)}, Lecture Notes in Computer Science, New York, USA, September
  2008. Springer Verlag.

\bibitem{miccai:07}
C.~Poupon, F.~Poupon, \underline{A.~Roche}, Y.~Cointepas, J.~Dubois, and J.-F. Mangin.
\newblock {Real-Time MR Diffusion Tensor and Q-Ball Imaging Using Kalman
  Filtering}.
\newblock In {\em Medical Image Computing and Computer-Assisted Intervention
  (MICCAI'07)}, volume 4791 of {\em Lecture Notes in Computer Science}, pages
  27--35, Brisbane, Australia, November 2007. Springer Verlag.

\bibitem{ipmi:07}
B.~Thirion, A.~Tucholka, M.~Keller, P.~Pinel, \underline{A.~Roche}, J.-F. Mangin, and J.-B.
  Poline.
\newblock {High Level Group Analysis of fMRI Data Based on Dirichlet Process
  Mixture Models}.
\newblock In {\em Information Processing in Medical Imaging (IPMI'07)}, Lecture
  Notes in Computer Science, pages 482--494, Kerkrade, Netherlands, 2007.
  Springer Verlag.

\bibitem{isbi:06}
S.~M\'eriaux, \underline{A.~Roche}, B.~Thirion, and G.~Dehaene-Lambertz.
\newblock Robust statistics for nonparametric group analysis in {fMRI}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'06)},
  pages 936--939, Washington, DC, USA, April 2006.

\bibitem{mmbia:06}
B.~Thirion, \underline{A.~Roche}, P.~Ciuciu, and J.-B. Poline.
\newblock {Improving sensitivity and reliability of fMRI group studies through
  high level combination of individual subjects results}.
\newblock In {\em Mathematical Methods in Biomedical Image Analysis
  (MMBIA'06)}, page~62, New York, USA, June 2006.

\bibitem{isbi:04b}
P.~Ciuciu, J.~Idier, \underline{A.~Roche}, and C.~Pallier.
\newblock Outlier detection for robust region-based estimation of the
  hemodynamic response function in event-related f{MRI}.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 392--395, Washington, DC, USA, April 2004.

\bibitem{isbi:04c}
E.~Duchesnay, \underline{A.~Roche}, D.~Rivi\`ere, D.~Papadopoulos-Orfanos, Y.~Cointepas,
  and J.-F. Mangin.
\newblock Population classification based on structural morphometry of cortical
  sulci.
\newblock In {\em International Symposium on Biomedical Imaging (ISBI'04)},
  pages 1276--1279, Washington, DC, USA, 2004.

\bibitem{miccai:02}
E.~Bardinet, P.~Cathier, \underline{A.~Roche}, and D.~Dormont.
\newblock {A Posteriori Validation of Pre-Operative Planning in Functional
  Neurosurgery By Quantification of Brain Pneumocephalus}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'02)}, volume 2488(1) of
  {\em Lecture Notes in Computer Science}, pages 323--330, Tokyo, Japan,
  November 2002. Springer Verlag.

\bibitem{miccai:01}
E.~Bardinet, A.C.F. Colchester, \underline{A.~Roche}, Y.~Zhu, Y.~He, S.~Ourselin,
  B.~Nailon, S.A. Hojjat, J.~Ironside, S.~Al-Sarraj, N.~Ayache, and J.~Wardlaw.
\newblock {Registration of Reconstructed Post Mortem Optical Data with MR Scans
  of the Same Patient}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 957--965, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miccai:01d}
H.~Delingette, E.~Bardinet, D.~Rey, J.-D. Lemarechal, J.~Montagnat,
  S.~Ourselin, \underline{A.~Roche}, D.~Dormont, J.~Yelnik, and N.~Ayache.
\newblock {YAV++: a Software Platform for Medical Image Processing and
  Visualization}.
\newblock In {\em Workshop on Interactive Medical Image Visualization and
  Analysis, satellite symposia of MICCAI'01}, Utrecht, Netherlands, October
  2001.

\bibitem{miccai:01c}
S.~Ourselin, E.~Bardinet, D.~Dormont, G.~Malandain, \underline{A.~Roche}, and N.~Ayache.
\newblock {Fusion of Histological Sections and MR Images: Towards the
  Construction of an Atlas of the Human Basal Ganglia}.
\newblock In W.J. Niessen and M.A. Viergever, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'01)}, volume 2208 of
  {\em Lecture Notes in Computer Science}, pages 743--751, Utrecht,
  Netherlands, October 2001. Springer Verlag.

\bibitem{miar:01}
X.~Pennec, N.~Ayache, \underline{A.~Roche}, and P.~Cachier.
\newblock {Non-rigid MR/US registration for tracking brain deformations}.
\newblock In {\em International Workshop on Medical Imaging and Augmented
  Reality (MIAR'01)}, pages 79--86, Shatin, Hong Kong, June 2001. IEEE Computer
  Society Press.

\bibitem{miccai:00c}
A.~Colchester, S.~Ourselin, Y.~Zhu, E.~Bardinet, Y.~He, \underline{A.~Roche}, S.~Al-Sarraj,
  B.~Nailon, J.~Ironside, and N.~Ayache.
\newblock {3-D Reconstruction of Macroscopic Optical Brain Slice Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 95--105, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{paris:00}
O.~Migneco, G.~Malandain, \underline{A.~Roche}, I.~Dygai, F.~Bussiere, N.~Ayache, and
  J.~Darcourt.
\newblock Absolute calibration of hmpao spect using xenon 133.
\newblock In {\em Congr\`es Europ\'een de M\'edecine Nucl\'eaire}, Paris,
  France, September 2000.

\bibitem{miccai:00b}
S.~Ourselin, \underline{A.~Roche}, S.~Prima, and N.~Ayache.
\newblock {Block Matching: a General Framework to Improve Robustness of Rigid
  Registration of Medical Images}.
\newblock In S.~Delp, A.M. DiGioia, and B.~Jaramaz, editors, {\em Medical Image
  Computing and Computer-Assisted Intervention (MICCAI'00)}, volume 1935 of
  {\em Lecture Notes in Computer Science}, pages 557--566, Pittsburgh, USA,
  October 2000. Springer Verlag.

\bibitem{wbir:99}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, and X.~Pennec.
\newblock {Automatic Alignment of Histological Sections}.
\newblock In F.~Pernus, S.~Kovacic, H.S. Stiehl, and M.A. Viergever, editors,
  {\em International Workshop on Biomedical Image Registration (WBIR'99)},
  pages 1--13, Bled, Slovenia, August 1999.


\subsection*{Abstracts}

\bibitem{adpd:17}
\underline{A.~Roche}, D.~Damian, F.~Pedron, B.~Mar\'echal, P.~Hagmann, P.~Maeder, R.~Meuli,
  T.~Kober, and J.-F. D\'emonet.
\newblock {Automated prediction of typical and mixed Alzheimer's disease using
  combined routine brain volumetry and cognitive assessment}.
\newblock In {\em Submitted to AD-PD conference}, Vienna, 2017.

\bibitem{ismrm:14}
\underline{A.~Roche}, D.~Schmitter, B.~Mar\'echal, D.~Ribes, A.~Abdulkadir, M.~Bach~Cuadra,
  A.~Daducci, C.~Granziera, S.~Kl\"oppel, P.~Maeder, R.~Meuli, and G.~Krueger.
\newblock {Volume-based vs. voxel-based brain morphometry in Alzheimer's
  disease prediction}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.
\newblock The first two authors contributed equally.

\bibitem{ohbm:05}
\underline{A.~Roche}.
\newblock {When is it a good time to smooth fMRI data?}
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'05) CD-Rom Neuroimage 26(1)}, Toronto, Canada, June 2005.

\bibitem{ohbm:04}
\underline{A.~Roche} and J.-B. Poline.
\newblock {Kalman Filtering for Real-Time fMRI Analysis}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:03}
\underline{A.~Roche}, F.~Kherif, G.~Flandin, and J.-B. Poline.
\newblock Should {fMRI} data be analyzed using a single {BOLD} response model
  across regions?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'03) CD-Rom Neuroimage 19(2)}, New York, USA, June 2003.

\bibitem{ismrm:16}
P.~Falkovskiy, B.~Mar\'echal, T.~Kober, P.~Maeder, R.~Meuli, J.-P. Thiran, and
  \underline{A.~Roche}.
\newblock {Impact of image acquisition systems on Alzheimer's disease-related
  atrophy detection}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16b}
P.~Falkovskiy, B.~Mar\'echal, S.~Yan, Z.~Jin, T.~Qian, K.~O'Brien, R.~Meuli,
  J.-P. Thiran, G.~Krueger, T.~Kober, and \underline{A.~Roche}.
\newblock {Quantitative comparison of MP2RAGE skull-stripping strategies}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ohbm:16}
P.~Falkovskiy, B.~Mar\'echal, P.~Maeder, T.~Kober, J.-P. Thiran, R.~Meuli, and
  \underline{A.~Roche}.
\newblock {Correcting Effects of Magnetic Resonance Field Strength on Brain
  Volumetry}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'16)}, Geneva, Switzerland, 2016.

\bibitem{ismrm:15}
G.~Bonnier, J.-P Thiran, G.~Krueger, C.~Granziera, and \underline{A.~Roche}.
\newblock {Concentration maps improve detection of gray matter alteration in
  cerebellum and deep gray matter structures}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:13}
K.~O'Brien, G.~Krueger, F.~Lazeyras, R.~Grueter, and \underline{A.~Roche}.
\newblock {A simple method to denoise MP2RAGE}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:11}
D.~Ribes, B.~Mortamet, M.~Bach~Cuadra, C.R. Jack, R.~Meuli, G.~Krueger, and
  \underline{A.~Roche}.
\newblock {Comparison of tissue classification models for automatic brain MR
  segmentation}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'11)}, Montreal, Canada, 2011.

\bibitem{ohbm:09}
L.~Favre, A.-L. Fouque, T.~Vincent, A.~Tucholka, M.~Keller, G.~Operto,
  B.~Thyreau, C.~Clouchoux, L.~Risser, A.~Moreno, D.~Geffroy, Y.~Cointepas,
  O.~Coulon, P.~Ciuciu, B.~Thirion, and \underline{A.~Roche}.
\newblock {A Comprehensive fMRI Processing Toolbox for BrainVISA}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'09)}, San Francisco, USA, 2009.

\bibitem{ismrm:16c}
S.~Haller, P.~Falkovskiy, R.~Meuli, J.-P. Thiran, G.~Krueger, K.-O. Lovblad,
  \underline{A.~Roche}, T.~Kober, and B.~Mar\'echal.
\newblock {Basic MR sequence parameters systematically bias automated brain
  volume estimation}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16d}
M.J. Fartaria, G.~Bonnier, T.~Kober, \underline{A.~Roche}, D.~Rotzinger, M.~Schluep,
  R.~Du~Pasquier, J.-P. Thiran, G.~Krueger, R.~Meuli, M.~Bach~Cuadra, and
  C.~Granziera.
\newblock {Longitudinal automated detection of white matter and cortical
  lesions in relapsing-remitting multiple sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:16e}
M.J. Fartaria, G.~Bonnier, T.~Kober, K.~O'Brien, \underline{A.~Roche}, B.~Mar\'echal,
  D.~Rotzinger, R.~Meuli, J.-P. Thiran, G.~Krueger, M.~Bach~Cuadra, and
  C.~Granziera.
\newblock {Focal cerebellar pathology in early relapsing-remitting multiple
  sclerosis patients: a MP2RAGE study at 3T and 7T MRI }.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'16)}, Singapore, 2016.

\bibitem{ismrm:15b}
M.J. Fartaria, G.~Bonnier, \underline{A.~Roche}, T.~Kober, R.~Meuli, D.~Rotinger,
  M.~Schluep, R.~Du~Pasquier, J.-P. Thiran, G.~Krueger, C.~Granziera, and
  M.~Bach~Cuadra.
\newblock {Exploration of advanced MR imaging contrasts for automated detection
  of white matter and cortical lesions in early-stages of multiple sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:15c}
Y.~Wang, B.~Mar\'echal, D.~Neumann, \underline{A.~Roche}, J.D. West, B.C. McDonald, M.A.
  Keiski, D.J. Smith, A.J. Saykin, and G.~Krueger.
\newblock {Detecting atrophy in chronic moderate and severe traumatic brain
  injury using an automated volume-based morphometry toolbox}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'15)}, Toronto, Canada, 2015.

\bibitem{ismrm:14b}
D.~Romascano, D.E. Meskaldji, G.~Bonnier, S.~Simioni, D.~Rotzinger, Y.-C. Lin,
  G.~Menegaz, \underline{A.~Roche}, M.~Schluep, and R.~Du~Pasquier.
\newblock {Cerebellar Connectomics Provide New Biomarkers in Early Multiple
  Sclerosis}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.

\bibitem{ismrm:14c}
G.~Bonnier, \underline{A.~Roche}, D.~Romascano, S.~Simioni, D.E. Meskaldji, D.~Rotzinger,
  Y.-C. Lin, G.~Menegaz, M.~Schluep, and R.~Du~Pasquier.
\newblock {Multiple Sclerosis Lesion Fingerprint Using Multicontrast MRI}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'14)}, Milan, Italy, 2014.

\bibitem{ismrm:13b}
M.~Bach~Cuadra, S.~Gelin, \underline{A.~Roche}, O.~Esteban, J.~Kober, T.~Marques,
  C.~Granziera, and G.~Krueger.
\newblock {Classical segmentation methods on novel MR imaging: a study of brain
  tissue segmentation of MP2RAGE vs MPRAGE}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:13c}
G.~Bonnier, T.~Stumpf, D.~Romascano, \underline{A.~Roche}, M.~Schluep, R.~Dupasquier,
  C.~Granziera, and G.~Krueger.
\newblock {Ultrafast T2 mapping of multiple sclerosis pathology in early
  disease}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'13)}, Salt Lake City, USA, 2013.

\bibitem{ismrm:12}
T.~Kober, \underline{A.~Roche}, D.~Ribes, R.~Meuli, and G.~Krueger.
\newblock {Atlas-free brain tissue segmentation using a single T1-weighted MRI
  acquisition}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'12)}, Melbourne, Australia, 2012.

\bibitem{ismrm:12b}
B.~Mar\'echal, T.~Kober, T.~Hilbert, D.~Ribes, N.~Chevrey, \underline{A.~Roche}, J.-P.
  Thiran, R.~Meuli, and G.~Krueger.
\newblock {Automated quality control in MR-based brain morphometry}.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'12)}, Melbourne, Australia, 2012.

\bibitem{ohbm:09b}
M.~Brett, J.~Taylor, C.~Burns, K.J. Millman, F.~Perez, \underline{A.~Roche}, B.~Thirion,
  and M.~D'Esposito.
\newblock {NIPY: an open library and development framework for FMRI data
  analysis}.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'09) Neuroimage 47:S196}, San Francisco, USA, 2009.

\bibitem{ismrm:08}
F.~Poupon, \underline{A.~Roche}, J.-F. Mangin, and C.~Poupon.
\newblock Real-time {MR} diffusion tensor and {Q-ball} imaging using {Kalman}
  filtering.
\newblock In {\em Annual Meeting of the International Society for Magnetic
  Resonance in Medicine (ISMRM'08)}, Toronto, Canada, 2008.

\bibitem{ohbm:06}
S.~M\'eriaux, \underline{A.~Roche}, G.~Dehaene-Lambertz, and J.-B. Poline.
\newblock When do mixed-effect models fail to improve detection sensitivity in
  f{MRI} group activation maps?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'06) CD-Rom Neuroimage 31(1)}, Florence, Italy, 2006.

\bibitem{ohbm:05b}
S.~M\'eriaux, \underline{A.~Roche}, J.~Catry, O.~Chauvin, and J.-B. Poline.
\newblock Robust statistics for f{MRI} group analyses.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'05) CD-Rom Neuroimage 26(1)}, Toronto, Canada, June 2005.

\bibitem{ohbm:04c}
E.~Duchesnay, \underline{A.~Roche}, D.~Rivi\`ere, D.~Papadopoulos-Orfanos, Y.~Cointepas,
  and J.-F. Mangin.
\newblock Guessing the sex from the shapes of cortical folds.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (OHBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:04d}
G.~Flandin, X.~Pennec, \underline{A.~Roche}, W.~Penny, N.~Ayache, and J.-B. Poline.
\newblock Multi-subject anatomo-functional classification for activation
  studies.
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{ohbm:04b}
S.~M\'eriaux, F.~Kherif, \underline{A.~Roche}, M.~Brett, L.~Garnero, and J.-B. Poline.
\newblock How frequently do we sample inhomogeneous group of subjects in f{MRI}
  studies?
\newblock In {\em Annual Meeting of the Organization for Human Brain Mapping
  (HBM'04) CD-Rom Neuroimage 22(1)}, Budapest, Hungary, June 2004.

\bibitem{cancero:99}
P.~Y. Bondiau, G.~Malandain, K.~Benezery, \underline{A.~Roche}, R.~Ferrand, D.~Ponvert,
  J.~L. Habrand, and J.~N. Bruneton.
\newblock {Fusions Scanner / IRM entri\`erement automatique: perspectives
  d'utilisation pour la radioth\'erapie}.
\newblock In {\em {Proc. du XIXe Forum de Canc\'erologie}}, Paris, June 1999.

\bibitem{rsna:99}
P.~Y. Bondiau, G.~Malandain, K.~Benezery, \underline{A.~Roche}, R.~Ferrand, D.~Ponvert,
  J.~L. Habrand, and J.~N. Bruneton.
\newblock {Automatic CT Scan and MRI Matching: Use for Brain Tumor
  Radiotherapy}.
\newblock In {\em {RSNA'99}}, Chicago, USA, November 1999. Published in
  Supplement to Radiology, 213:236.

\bibitem{dijon:99}
S.~Ourselin, C.~Sattonnet, \underline{A.~Roche}, and G.~Subsol.
\newblock {Automatic alignment of histological sections for 3D reconstruction
  and analysis}.
\newblock In {\em 6\`eme congr\`es annuel de l'Association Fran\c caise de
  Cytom\'etrie}, Dijon, France, October 1999. Published in the Analytical
  Cellular Pathology Journal, 19(3):123.


\subsection*{Preprints}

\bibitem{rr:15}
\underline{A.~Roche}.
\newblock {Composite Bayesian inference}.
\newblock Technical report, 2015.
\newblock Available at arxiv.org (arXiv:1512.07678).

\bibitem{rr:12}
\underline{A.~Roche}.
\newblock {EM algorithm and variants: an informal tutorial}.
\newblock Technical report, 2012.
\newblock Available at arxiv.org (arXiv:1105.1476).

\bibitem{rr:00}
\underline{A.~Roche}, X.~Pennec, G.~Malandain, N.~Ayache, and S.~Ourselin.
\newblock {Generalized Correlation Ratio for Rigid Registration of 3D
  Ultrasound with MR Images}.
\newblock Technical Report 3980, INRIA, July 2000.

\bibitem{rr:99}
\underline{A.~Roche}, G.~Malandain, and N.~Ayache.
\newblock {Unifying Maximum Likelihood Approaches in Medical Image
  Registration}.
\newblock Technical Report 3741, INRIA, July 1999.

\bibitem{rr:98}
\underline{A.~Roche}, G.~Malandain, N.~Ayache, and X.~Pennec.
\newblock {Multimodal Image Registration by Maximization of the Correlation
  Ratio}.
\newblock Technical Report 3378, INRIA, August 1998.

\bibitem{rr:01}
S.~Granger, X.~Pennec, and \underline{A.~Roche}.
\newblock {Rigid Point-Surface Registration using Oriented Points and an EM
  Variant of ICP for Computer Guided Oral Implantology}.
\newblock Technical Report 4169, INRIA, April 2001.

\bibitem{rr:99b}
A.~Guimond, \underline{A.~Roche}, N.~Ayache, and J.~Meunier.
\newblock {Multimodal Brain Warping Using the Demons Algorithm and Adaptative
  Intensity Corrections}.
\newblock Technical Report 3796, INRIA, November 1999.

\bibitem{rr:98b}
S.~Ourselin, \underline{A.~Roche}, G.~Subsol, X.~Pennec, and C.~Sattonet.
\newblock {Automatic Alignment of Histological Sections for 3D Reconstruction
  and Analysis}.
\newblock Technical Report 3595, INRIA, December 1998.


\subsection*{Patents}

\bibitem{patentapp:12}
\underline{A.~Roche}, T.~Kober, and G.~Krueger.
\newblock Automated atlas-free brain tissue classification using specific
  magnetic resonance imaging sequences, 2013.
\newblock Owner: Siemens Schweiz AG, Switzerland. US patent application
  13/855972.

\bibitem{patent:05}
\underline{A.~Roche}, M.~Brady, J.~Declerck, and O.~Corolleur.
\newblock Similarity measures.
\newblock United States patent US2005152617, July 2005.
\newblock Owner: Mirada Solutions Ltd, UK. Also published as EP1513104
  (Europe), JP2005169077 (Japan).

\bibitem{patent:03}
\underline{A.~Roche}, N.~Ayache, G.~Malandain, and X.~Pennec.
\newblock Electronic device for automatic registration of images.
\newblock United States patent US6539127, March 2003.
\newblock Owner: INRIA, France. Also published as EP0977148 (Europe), FR2781906
  (France).

\bibitem{patent:15}
K.~O'Brien and \underline{A.~Roche}.
\newblock {A simple method to denoise MP2RAGE}.
\newblock United States patent US2013P01055, 2013.
\newblock Owner: Siemens Schweiz AG, Switzerland. Also published as EP 2765437.

\bibitem{patent:12}
C.~Poupon, F.~Poupon, \underline{A.~Roche}, and J.-F. Mangin.
\newblock Real-time magnetic resonance diffusion imaging.
\newblock United States patent US8330462, 2007.
\newblock Owner: CEA, France. Also published as EP2203758A1 (Europe).


\subsection*{Theses}

\bibitem{these:01}
\underline{A.~Roche}.
\newblock {\em Recalage d'images m\'edicales par inf\'erence statistique}.
\newblock PhD thesis, Universit\'e de Nice-Sophia Antipolis, February 2001.

\bibitem{dea:96}
\underline{A.~Roche}.
\newblock {Mod\'elisation de la perception du tempo par oscillateurs
  adaptatifs}.
\newblock Master's thesis, {Universit\'e Pierre et Marie Curie, Paris VI},
  1996.

\end{thebibliography}
